---
title: "Quiz1_yourStudent_ID"
author: "Add Your Name"
date: "2023-12-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1 - Multiple Choice Questions

Put your answers for MC questions in the table below;

MC   answer
---  ---
(a)           
(b)   
(c)   
(d)   
(e)    



### Question 2

```{r, echo = TRUE}
sample_x <- c(78, 86, 97, 91, 83, 89, 92, 88, 79, 68)
sample_x
```

#### part a)

```{r, echo = TRUE}
set.seed(1413) #DO NOT REMOVE THIS

#your code here
```

#### part b)


```{r, echo = TRUE}
# add your code here
```

#### part c) 

```{r, echo = TRUE}
# add your code here
```


#### Question 3 

Run this code before continue. note that the data is saved under the name `mydata`. 

```{r, echo= TRUE}
#install.package("mlbench") if needed
library(mlbench)
data("PimaIndiansDiabetes2", package = "mlbench")
mydata <- PimaIndiansDiabetes2
```

#### part a) 

Write your answer here

#### part b) 

```{r, echo= TRUE}
#your code
```

#### part c)
```{r, echo= TRUE}
# your codes here
```

#### part d)

Add your explanation here...

#### part e) 

```{r, echo= TRUE}
# here your code
```

Add your explanation here: 


#### part f) 

```{r, echo= TRUE}
#your code
```
Add your explanation here: 




### Question 4:


Run this code before continue. note that the data is saved under the name `mydata`. 

```{r, echo= TRUE}
attach(mtcars)
# you can select a column of a dataset by $ operator, for example
# wt_col <- mtcars$wt
```

#### part a

Add your answer here.

#### part b

```{r, echo= TRUE}
#your code
```

#### part c

```{r, echo= TRUE}
#your code
```