---
title: "Quiz_1_32376881"
author: "Dhun Sheth"
date: "2023-12-04"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1 - Multiple Choice Questions

Put your answers for MC questions in the table below;

MC   answer
---  ---
(a)   iii        
(b)   iii
(c)   ii
(d)   iv
(e)   i, iv



### Question 2

```{r, echo = TRUE}
sample_x <- c(78, 86, 97, 91, 83, 89, 92, 88, 79, 68)
sample_x
```

#### part a)

```{r, echo = TRUE}
set.seed(1413) #DO NOT REMOVE THIS

sample_x_sorted <- sort(sample_x)

print(mean(sample_x_sorted, trim=0.2))

```

#### part b)


```{r, echo = TRUE}
# add your code here

x1000 <- replicate(1000, mean(sample(sample_x_sorted, size = 10, replace= TRUE)
                              , trim=0.2))
SE <- sd(x1000)
print(SE)
```

#### part c) 

```{r, echo = TRUE}
# add your code here

# 90% CI
lower_bound <- quantile(x1000,0.05)
upper_bound <- quantile(x1000,0.95)
CI <- c(lower_bound, upper_bound)
print(CI)
## OR

ci <- mean(sample_x, trim=0.2) + c(-qnorm(1-0.05), qnorm(1-0.05))*SE
print(ci)
```


#### Question 3 

Run this code before continue. note that the data is saved under the name `mydata`. 

```{r, echo= TRUE}
#install.package("mlbench") if needed
library(mlbench)
data("PimaIndiansDiabetes2", package = "mlbench")
mydata <- PimaIndiansDiabetes2
```

#### part a) 

Because we want to predict if the patient has diabetes or not (1 for yes, and 0 
for no), we can use the binary logistic regression using the logit link function. 

#### part b) 

```{r, echo= TRUE}
#your code
str(na.omit(mydata))
new_mydata <- na.omit(mydata)
```

#### part c)
```{r, echo= TRUE}
# your codes here

glm.reg <- glm(diabetes~., data=new_mydata, family=binomial)

```

#### part d)

Coefficient for glucose is `r glm.reg$coefficients[[3]]`, because it is positive, there is a positive
relationship between glucose and predicting diabetes, so the higher glucose is
the more likely they are to have diabetes. 

#### part e) 

```{r, echo= TRUE}
# here your code
print(summary(glm.reg))
```

Add your explanation here: glucose, mass, and pedigree are below the 0.05 
significance level and are thus significant.


#### part f) 

```{r, echo= TRUE}
#your code
step(glm.reg, direction = 'backward')
```
Add your explanation here: The final model chosen based on backwards selection
includes pregnant, glucose, mass, pedigree, and age predictor variables.




### Question 4:


Run this code before continue. note that the data is saved under the name `mydata`. 

```{r, echo= TRUE}
attach(mtcars)
# you can select a column of a dataset by $ operator, for example
# wt_col <- mtcars$wt
data <- mtcars[,c(1,4,6)]
```

#### part a

assuming there is an intercept for the model, design matrix (X) would be 32x3 (nxp).

#### part b

```{r, echo= TRUE}
#your code

n <- nrow(data) # number of rows (n) of design matrix
p <- ncol(data[-1]) + 1 # number of columns (p) of design matrix -- remove y column and add 1's col for intercept
X <- data[,-1] # remove "y" column to form model matrix
X1 <- rep(1, n) # column for intercept
X <- cbind(X1, X) # append 1's to create full model matrix
y <- data[, 1]

QR <- qr(X) # QR decomposition of X
Q <- qr.Q(QR, complete=TRUE) # complete=TRUE gives Q1 & Q2
R <- qr.R(QR, complete=TRUE)
Q1 <- Q[, 1:p]
Q1y <- t(Q1)%*%y
U <- qr.R(QR) # if you want all of R, use complete=TRUE

# parameter estimate
betahat <- solve(U, Q1y)
print(betahat)

```

#### part c

```{r, echo= TRUE}
#your code

Q2 <- Q[, -(1:p)]
Q2y <- t(Q2)%*%y
SSE <- t(Q2y)%*%Q2y
print(SSE)

MSE <- SSE/(n-p)
```

#### Part d

```{r, echo= TRUE}
Cii <- sqrt(diag(solve(t(U)%*%U)))
SEii <- Cii*as.numeric(sqrt(MSE))
print(SEii)

```