---
title: "DATA-581 Quiz 2 Prep"
author: "Dhun Sheth"
date: "2023-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(HMM)
```

## Simulating 1000 movements of the mouse (MC example - Lecture 5)
```{r}

Ntransitions <- 1000 # number of transitions
P <- matrix(c(0, 1, 0, 0,
              0.5, 0, 0.5, 0,
              0, 0.5, 0, 0.5,
              0, 0, 1, 0), nrow = 4, byrow = TRUE) # P is the transition matrix
X <- numeric(Ntransitions) #initializing the chain
current.state <- 1 # initial state X_0 = 1
for (t in 1:Ntransitions) {
  current.state <- sample(1:4, size = 1, prob = P[current.state, ])
  X[t] <- current.state
}

table(X) # counts in each state
table(X)/Ntransitions # probability being in each state after 1000 moves

```

## Calculate stationary matrix (Lecture 6)
```{r}
P33<- matrix(c(0, 0.4, 0.6,
               0.5, 0, 0.5,
               0.25, 0, 0.75), nrow = 3, byrow = TRUE)

A <- t(P33) - diag(rep(1,3))
A <- rbind(A, rep(1,3))
RHS <- c(rep(0,3), 1)
qr.solve(A, RHS) # no longer a square system


```


## Simulating from an infinite MC (time-reversible) (lec 6 slide 28/29)
```{r}

pi.fun <- function(i) {
  out <- 0
  if (i > 0) out <- 1/(i+1)^4
  out
}

N <- 20000
X <- numeric(N)
current.state <- 50 # initialize the Markov chain
for (n in 1:N) {
  i <- current.state
  P <- c(min(pi.fun(i-2)/pi.fun(i), 1), 
         min(pi.fun(i-1)/pi.fun(i), 1),
         min(pi.fun(i+1)/pi.fun(i), 1),
         min(pi.fun(i+2)/pi.fun(i), 1))/6
  P0 <- 1 - sum(P)
  P <- c(P[1:2], P0, P[3:4])
  transition <- sample(seq(-2,2,1), size = 1, prob = P)
  current.state <- current.state + transition
  X[n] <- current.state
}
n <-1000
observedDist <- table(X[-c(1:n)]) # removing first 1000 observations "burn-in period"
pi <- observedDist/(N-n)
pi

```

## Given some distribution for pi (slide 37)
```{r}

posterior.fun <- function(i, x) {
  out <- 0
  if (i >= x) {
    out <- dpois(i, lambda = 20) * dbinom(x, size = i, prob = .5)
    }
  out
}

pi.fun <- function(i) {
  posterior.fun(i, x=5)
}

Ntransitions <- 20000
X <- numeric(Ntransitions)
current.state <- 40 # initialize the Markov chain
for (n in 1:Ntransitions) {
  i <- current.state
  P <- c(min(pi.fun(i-2)/pi.fun(i), 1),
         min(pi.fun(i-1)/pi.fun(i), 1),
         min(pi.fun(i+1)/pi.fun(i), 1),
         min(pi.fun(i+2)/pi.fun(i), 1))/6
  P0 <- 1 - sum(P)
  P <- c(P[1:2], P0, P[3:4])
  transition <- sample(seq(-2,2,1), size = 1, prob = P)
  current.state <- current.state + transition
  X[n] <- current.state
}
observedDist <- table(X[-c(1:1000)])
pi <- observedDist/(N-1000)

par(mfrow=c(1,2))
theoryDist <- 20000*dpois(0:30, lambda = 20)
names(theoryDist) <- 0:30
barplot(theoryDist, main = "Prior")
barplot(observedDist, main = "Posterior")


```


## HMM Package (HMM package and Viterbi path)
```{r}

hmm <- initHMM(c("0","1"), c("0","1"), 
               transProbs=matrix(c(1, 2, 2, 1)/3,2),
               emissionProbs=matrix(c(3, 1, 1, 3)/4, 2))

observations <- as.character(c(1, 0, 0, 0, 1, 1, 0))

viterbi <- viterbi(hmm, observations)
print(viterbi)

```

## [HMM] More on Viterbi and training Viterbi based on an initial guess of the transition and emission matrix (when transition and emission matrix are not known)
```{r}

P <- matrix(c(.1, .9, .2, .8), nrow=2, byrow=TRUE)
E <- matrix(c(.75, .2, 0.05, .05, 0.4, .55), nrow=2, byrow=TRUE)
current.state <- 1; n <- 500
X <- numeric(n)
Y <- numeric(n)
for (i in 1:n) {
  current.state <- sample(0:1, size = 1, prob = P[current.state+1,])
  X[i] <- current.state #hidden states
  Y[i] <- sample(0:2, size = 1, prob = E[current.state+1,])
  #Y observed data
}
Y[1:5] # first few simulations of Y

hmm <- initHMM(c("0","1"), c("0","1", "2"),
               transProbs=matrix(c(.5, .5, .5, .5),2),
               emissionProbs=matrix(c(.5, .4, .1, .8, .05, .15), 2))
observations <- as.character(Y)
simDataFit <- viterbiTraining(hmm, observations)
print(simDataFit$hmm$transProbs)
print(simDataFit$hmm$emissionProbs)

```