---
title: "DATA-580 Lab 1"
author: "Dhun Sheth"
date: "2023-09-14"
geometry: margin=2cm
output: pdf_document
---

```{r setup, include=FALSE}
rng.chisq <- function(x) {
  # x is output from a discreat uniform pseudorandom number
  n=length(x)
  x=as.vector(table(x))#Oi
  m=length(x)
  p <- rep(1,m)/m
  Ex <- n*p#Ei
  chisq <- sum((x-Ex)^2/Ex)
  pvalue <- 1-pchisq(chisq, m-1)
  list(test.statistic=chisq, p.value=pvalue, df=m-1)
}
```

## Question 1
```{r question_1}
bpm_data=readRDS(file = "BPM.rds")
```

### Part A
```{r part_1a}
rng.chisq(bpm_data[1:150,2])[2]
```
The p-value is greater than 0.05, and not very small, there is little evidence against the uniformity hypothesis
and thus the BPM in the first phase follows a uniform distribution. 

### Part B
```{r part_1b}
rng.chisq(bpm_data[,2])[2]
```
The p-value is less than 0.05 and is very small, there is large evidence against the uniformity hypothesis and thus the entire sequence of BPM does not follow a uniform distribution.

<p>&nbsp;</p>

## Question 2
```{r question_2}
exchange_data=readRDS(file = "Exchanges.rds")
```

### Part A
```{r lag_plot}
lag.plot(exchange_data[,4], lags=5)
acf(exchange_data[,4],5)
```

### Part B
Based on the plot, there are several spikes in the plot which exceed the blue line which indicates dependence. In addition, the lag plots also show a linear dependence. 

<p>&nbsp;</p>

## Question 3
```{r question_3}
dna_data=read.table("DNA.txt", header = FALSE)
```

### Part A
```{r part_3a, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
list_C1_AT = c()
for (k in dna_data[,2]){
  n=nchar(k)
  i=1
  x=rep(0,n)
  while(i<(n-1)){
    if(substring(k,i,i)=="A") x[i]=1
    if(substring(k,i+1,i+1)=="T") x[i+1]=1
    i=i+2
  }
  sigma2=sum(x^2)/n-(mean(x))^2
  C1=(sum(x[1:(n-1)]*x[2:n])/(n-1)-mean(x[1:(n-1)])*mean(x[2:n]))/sigma2
  list_C1_AT = append(list_C1_AT,C1)
}
list_C1_AT[1:5] # printing seperatly so numbers don't get cut off
list_C1_AT[6:length(list_C1_AT)] # printing seperatly so numbers don't get cut off
```
### Part B
```{r part_3b, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
list_C1_AC = c()
for (k in dna_data[,2]){
  n=nchar(k)
  i=1
  x=rep(0,n)
  while(i<(n-1)){
    if(substring(k,i,i)=="A") x[i]=1
    if(substring(k,i+1,i+1)=="C") x[i+1]=1
    i=i+2
  }
  sigma2=sum(x^2)/n-(mean(x))^2
  C1=(sum(x[1:(n-1)]*x[2:n])/(n-1)-mean(x[1:(n-1)])*mean(x[2:n]))/sigma2
  list_C1_AC = append(list_C1_AC,C1)
}
list_C1_AC[1:5] # printing seperatly so numbers don't get cut off
list_C1_AC[6:length(list_C1_AC)] # printing seperatly so numbers don't get cut off
```

