---
title: "Data-580 Lab 3"
author: "Dhun Sheth"
date: "2023-09-25"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

### Part A

$$F(x)=log(x) => U=log(x) => x=e^U$$

### Part B

```{R question_1b}

U <- runif(100000)
x <- exp(U)

```


### Part C

```{R question_1c}

mean_x <- mean(x)
var_x <- var(x)
prob_x_less_than_2 <- mean(x < 2)

```

$E[x]=$ `r mean_x` \
$Var[x]=$ `r var_x` \
$P(x < 2)=$ `r prob_x_less_than_2` \


### Part D

$$pdf=\frac{d}{dx}cdf = \frac{d}{dx}log(x)=\frac{1}{x}=f(x)$$
$$E[x]=\int_a^{b}x*f(x)dx = \int_1^{e}1~dx=[x]_1^e=e-1=1.7182818$$
`r exp(1)-1`

\

$$E[x^2]=\int_a^{b}x^2*f(x)dx = \int_1^{e}x~dx=[\frac{x^2}{2}]_1^e=\frac{e^2-1}{2}=3.194528$$
`r (exp(1)**2-1)/2`

$$Var[x]=E[x^2]-E[x]^2=3.194528 ~- 1.7182818^2= 0.2420357$$

`r 3.194528-1.7182818**2` \
\
Comparing to the mean and variance estimate from Part C, we can see the values are very close. 
\

## Question 2

### Part A
```{r question_2a}

curve(dgamma(x, shape = 0.5, scale = 1), 0, 5)
curve(dgamma(x, shape = 1, scale = 1), 0, 5, add=TRUE, col=2)
curve(dgamma(x, shape = 3, scale = 1), 0, 5, add = TRUE, col=3)

```
\

Density curve with shape=0.5 and scale=1 is decreasing to 0 as x goes to infinity, and approached infinity as x goes to 0. \

\

Density curve with shape=1 and scale=1 is decreasing to 0 as x goes to infinity, and approaches 1 as x goes to 0. \

\ 

Density curve with shape=3 and scale=1 starts at 0,0 and increases, peaking at around x=2 with dgamma highest around 0.4, before decreasing to 0 as x goes to infinity. 


### Part B

```{r question_2b}

prob_3 <- pgamma(3, shape=5, scale=2)
prob_7 <- pgamma(7, shape=5, scale=2)

```

$P(x<=3)=$ `r prob_3` \
$P(x<=7)=$ `r prob_7`


### Part C
```{r question_2c}

x <- rgamma(100000, shape=5, scale=2)
sim_prob_3<-mean(x<=3)
sim_prob_7<-mean(x<=7)

```

Simulated and theoretical values are close to each other. \

Simulated Values: \
$P(x<=3)=$ `r sim_prob_3` \
$P(x<=7)=$ `r sim_prob_7` \

Theoretical Values: \
$P(x<=3)=$ `r prob_3` \
$P(x<=7)=$ `r prob_7` \

## Question 3

### Part A
```{r question_3a}

prob_norm_3 <- pnorm(3, mean=5, sd=2)
prob_norm_7 <- pnorm(7, mean=5, sd=2)

```

$P(x<=3)=$ `r prob_norm_3` \
$P(x<=7)=$ `r prob_norm_7`

### Part B
```{r question_3b}


x_norm <- rnorm(10000, mean=5, sd=2)
sim_norm_prob_3<-mean(x_norm<=3)
sim_norm_prob_7<-mean(x_norm<=7)

```

Simulated and theoretical values are very close to each other. \

Simulated Values: \
$P(x<=3)=$ `r sim_norm_prob_3` \
$P(x<=7)=$ `r sim_norm_prob_7` \

Theoretical Values: \
$P(x<=3)=$ `r prob_norm_3` \
$P(x<=7)=$ `r prob_norm_7` \

### Part C

### Part i., ii., iii.
```{r question_3c}

z_norm <- rnorm(10000)
y <- 2*x_norm + z_norm

plot(x_norm,y)
abline(0,2)
```
\

The (x,y) scatter plot forms a linear pattern which is concentrated around (5,10) and goes down in density as x moves away from 5. 

Because x_norm has a standard deviation of 2, and z has a standard deviation of 1, I would expect y = 2x, and so I would expect abline(0,2) which is the line y(x)=2x to pass through the middle of scatter plot (x,y)....and this is exactly what happens. 

### Part D
```{r question_3d}

z_norm_changed <- rnorm(10000, mean=0, sd=0.5)
y_part_d <- 2*x_norm + z_norm_changed

plot(x_norm,y_part_d)
abline(0,2)
```

### Part E
The scatter plot with z having a standard deviation of 0.5 becomes more closer to the line y(x)=2x. \
\

As the standard deviation of z gets larger than 1, the standard deviation of y will also get larger and the scatter plot will move away from a linear pattern matching the line y(x)=2x, and as the standard deviation of z gets smaller and closer to 0, the scatter plot will converge to the line y(x)=2x.