---
title: "DATA-581 Lab 2"
author: "Dhun Sheth"
date: "2023-11-29"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(MASS)
knitr::opts_chunk$set(echo = TRUE)
library(MPV)
```

## Question 3 (Exercise 5 from Ch. 4)
### Part A
```{r question_3_part_a}

y <- c(1,3,5,6,5,7)
x <- c(0,1,2,3,4,5)
X <- matrix(c(1,1,1,1,1,1,0,1,2,3,4,5), nrow=6)
print(X)

```

### Part B
```{r question_3_part_b}

X.QR <- qr(X)
R <- qr.R(X.QR, complete=TRUE)
Q <- qr.Q(X.QR, complete=TRUE)
print(Q)
print(R)
print(Q%*%R)
```

### Part C
```{r question_3_part_c}

U <- qr.R(X.QR)
U_inv <- solve(U) 
print(U_inv)

```
### Part D
```{r question_3_part_d}

p <- ncol(R)
n <- nrow(X)
Q1 <- Q[,1:p]

beta <- U_inv %*% t(Q1) %*% y
print(beta)

```

### Part E
```{r question_3_part_e}

y_pred <- beta[1] + beta[2]*x
QR_RSS <- sum((y_pred-y)^2)
print(QR_RSS)

```

### Part F
```{r question_3_part_f}

error_var <- QR_RSS/(n-p)
print(error_var)

```

### Part G
```{r question_3_part_g}

SE_beta <- sqrt(error_var)*sqrt(diag(solve(t(U)%*%U)))
t_statistic <- beta[2]/SE_beta[2]
p_value <- 2*(1-pt(t_statistic,n-p))
print(p_value)
```

### Part H
The test statistic follows a t-distribution on n-p degrees of freedom, 6-2 = 4 D.o.F. The p-value of the t-test is `r p_value` and because it is below the 0.05 significant level, there is enough evidence to reject the null hypothesis.


## Question 5 (Exercise 2 from Ch. 8)
### Part A
```{r question_5_part_a}

glm_reg <- glm(y~x, data=p13.2, family=binomial())
summary(glm_reg)

```

\(log(\frac{p}{1-p}) = `r glm_reg$coefficients[1]` + `r glm_reg$coefficients[2]`*x\)

\(p = \frac{e^{`r glm_reg$coefficients[1]` + `r glm_reg$coefficients[2]`*x}}{1 + e^{`r glm_reg$coefficients[1]` + `r glm_reg$coefficients[2]`*x}}\)

### Part B
```{r question_5_part_b}

y.logit2 <- glm(y~x, data = p13.2, family = binomial(link = "logit"))
anova(y.logit2, glm_reg)

plot(y~x, data=p13.2)
x_value <- 35000:75000
lines(x_value, predict(glm_reg,
newdata=data.frame(x = x_value), type = "response"))

```

There is only 1 predictor variable and the residual deviance is not much larger than the degree's of freedom, hence, the model is acceptable.

In addition, looking at the graph and estimated curve, the model seems acceptable.

### Part C
\(p= `r exp(glm_reg$coefficients[1] + glm_reg$coefficients[2]*40000)/(1+exp(glm_reg$coefficients[1] + glm_reg$coefficients[2]*40000))` => \) hence unlikely they own their home.  

## Question 6 
### Part A, B
```{r question_6_part_a}

epil_reg <- glm(y~age+trt, data=epil, family=poisson)
print(summary(epil_reg))

```

Based on the p-values being below the 0.05 significance level, all coefficients (intercept, age, trt) are significant. 

### Part C 
```{r question_6_part_c}

print(confint(epil_reg))
```

### Part D
The beta for treatment is `r epil_reg$coefficients[3]`. Because it is negative, this shows a negative relationship indicating a reduction in seizures with treatment. In addition, the treatment coefficient is significant. 

### Part E
```{r question_6_part_e}

new_data <- data.frame(age = 20, trt = "progabide")
pred_y <- predict(epil_reg, newdata = new_data, type = "response")
print(pred_y)

```

Predicted number of seizures for a 20-year-old with treatment: " `r round(pred_y)`